One-dimensional Classical Gas in a Harmonic Trap with Repulsive Interaction

### 2017-12-09 draft_06.tex
working on thermalization section, changed some figures

With interactions, a term $F_0(1-|r|)\theta(1-|r|)$ is added, where $r$ is the relative
displacement; as shown in Figure \ref{fig:thermalization2}.  Now, the relative motion is no longer
harmonic; one could regard the resulting motion has having a slightly different frequency, or a
collection of frequencies whose center is shifted slightly from $\omega_0$.  Since the
center-of-mass motion is still of frequency $\omega_0$, we have a superposition of slightly
different frequencies, resulting in beating dynamics.  This is clearly seen in the time evolution of
individual energies shown in Figure \ref{fig:thermalization2}.  The two panels correspond to
different internal energy $E_i$ (defined as the total energy minus the center-of-mass energy).  For
larger $E_i$, the distortion of the effective potential (at constant $F_0$) plays a smaller role in
shifting the effective frequency of relative motion; hence the beat frequency is smaller.

This illustrates a simple mechanism hindering the redistribution of energy between particles, which
is necessary for thermalization or relaxation.  As seen in the example dynamics shown, the
difference in energy between the two particles is sustained over time.  Of course thermalization is
not expected anyway in a two-particle system, but we will see below how this basic effect continues
to play a role for larger $N$.

-------

%% \begin{figure}[h]
%% \subfigure[$E_i=2$]{
%% \begin{minipage}[b]{0.22\textwidth}
%% \includegraphics[width=1\textwidth]{ZhiyuPictures/10_23_N=2_F0=1_Er=2_single_particle_energy_600_500_Font24.eps} 
%% %\includegraphics[scale=0.2]{ZhiyuPictures/twoparticle_Er=2_phasespace.eps}
%% \end{minipage}
%% }
%% \subfigure[$E_i=3$]{
%% \begin{minipage}[b]{0.22\textwidth}
%% \includegraphics[width=1\textwidth]{ZhiyuPictures/10_23_N=2_F0=1_Er=3_single_particle_energy_600_500_Font24.eps} 
%% %\includegraphics[scale=0.2]{ZhiyuPictures/twoparticle_Er=3_phasespace.eps}
%% \end{minipage}
%% }
%% \subfigure[$E_i=5$]{
%% \begin{minipage}[b]{0.22\textwidth}
%% \includegraphics[width=1\textwidth]{ZhiyuPictures/10_23_N=2_F0=1_Er=5_single_particle_energy_600_500_Font24.eps} 
%% %\includegraphics[scale=0.2]{ZhiyuPictures/twoparticle_Er=5_phasespace.eps}
%% \end{minipage}
%% }
%% \subfigure[$E_i=10$]{
%% \begin{minipage}[b]{0.22\textwidth}
%% \includegraphics[width=1\textwidth]{ZhiyuPictures/10_23_N=2_F0=1_Er=10_single_particle_energy_600_500_Font24.eps} 
%% %\includegraphics[scale=0.2]{ZhiyuPictures/twoparticle_Er=10_phasespace.eps}
%% \end{minipage}
%% }

%% \caption{the dependence of beat frequency on internal energy. The diagram is the energy of two
%%   particles. Here, $F_0$ is set to be 1. In this case, the larger the $E_i$, the lower the beat
%%   frequency. \label{fig:thermalization1}}
%% %Actually in very low energy, we can retrieve a beat, but that is for different reason -- it is a
%% %beat between $\omega_0$ and $\frac{\omega_0}{2}+\delta $
%% %
%% \end{figure}

@@

Now let's consider the three-particle case. For two particle, the motion is non-chaotic. On the other hand, intuitively, we will say that three-body motion is chaotic so that the system could be ``thermalized" soon. However, in some cases, the time scale of thermalizing could still be very long. Suppose two of them, say, A and B, has small internal energy, which means their distance and relative velocity are both small. Meanwhile, suppose particle C has some energy quite different from A \& B. In this case, A \& B will often be in the interaction, while C will pass both of them at a high speed in each period. How will energy transfer between C and the two-particle system A and B? Since the relative velocity of C and the two-particle system is usually large, C will pass A-B pair in a short time $\tau\sim\frac{\sigma}{R}$ ($\tau\ll\frac{2\pi}{\omega}$). C gives A a push when approaching it, and then push A back when leaving.  As has been discussed in the two-particle case, this process is equivalent to giving A a very small velocity ($\sim O(\frac{\sigma}{R})$) in the background of a trap. Since both the position and velocity of A and B are close($|x_A-x_B|\sim\sigma$), the velocity increase of A \& B are almost the same (difference$\sim O(\frac{\sigma^2}{R^2})$). In this manner, the passing of particle C only kicks the center of mass of the A-B pair slightly, leaving the internal motion of the pair unaffected. In another word, the existence of C will not have significant effect on the energy transfer between A and B, but only ``dance" with their center of mass. The physics of the ``dance" is similar to the dance between two particle. As is shown in Fig.\ref{fig:thermalization4}, the particles with energy close to each other tends to form a pair with small internal energy and the pair's internal energy transfer is relatively stable.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}
\centering
\includegraphics[scale=0.5]{ZhiyuPictures/pair1_pre.eps}
\caption{Time evolution of single-particle energy}
\label{fig:thermalization4}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

The argument above still holds in many-particle case. Once we start from some configuration where n particles have a set of close energy levels $\left\lbrace E_n\right\rbrace $ and another m particles have another set of energy levels close to each other $\left\lbrace E_m\right\rbrace $, and $\left\lbrace E_n\right\rbrace $ is quite different from $\left\lbrace E_n\right\rbrace $, we will find these two systems oscillating on the energy level diagram at low frequency. 
The discussion above gives us some pictures about the non-ergodic state. For these state, life time is rather long so that once the system reach such configuration (or certain energy distribution), it will take a long time(more than $10^4$ periods for N=5 case) to decay.

\subsection{Thermalization condition}
The main obstruct to get an thermalized state is the low frequency beat mentioned before. As a result, to achieve ergodic state, we have to avoid such low frequency oscillation. According to our discussion before, the solution is to make internal energy of each two-particle pair not ``too large". Though it is impossible to express the internal energy of every pair in terms of the total energy $E$, we can estimate it by the average energy, which is $\frac{E}{N}$. At least they are of the same order. The thermalization condition could be given by:
\begin{equation}
F_0\sigma\sim\frac{E}{N}
\label{eq:thermalizatiton condition}
\end{equation}

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h]
\centering
%\includegraphics[scale=0.2]{ZhiyuPictures/distribution1_scanF0_E=100_3.eps}
%\caption{Distribution scan $F_0$, $E=100$, $N=5$, $\sigma=1$ {\color{red}{Run again!}} }
\includegraphics[width=0.32\textwidth]{ZhiyuPictures/N=5_energydistribution_500_400_Font16.eps} 
\includegraphics[width=0.32\textwidth]{ZhiyuPictures/N=10_energydistribution_500_400_Font16.eps}
\includegraphics[width=0.32\textwidth]{ZhiyuPictures/N=20_energydistribution_500_400_Font16.eps}
\caption{the energy distribution measured in $\sim 10000$ time units, as will be verified later, Green curves corresponds to Boltzmann distribution; Red curves are obviously non-Boltzmann, while the yellow ones are intermediate case, represent ``thermalization threshold". In three figures for different N, we choose E/N identical (=1000), and thermalization is reached at $F_0=500\sim1000$, which is consistent with our prediction in Eq.\ref{eq:thermalizatiton condition}}
\label{fig:thermalization5}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[h]
\subfigure[N=10, $F_0=200$]{
\begin{minipage}[b]{0.4\linewidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/N=10_single_particle_energy_F0=200_s_400_300_Font12.eps}
\label{fig:thermalization6}
\end{minipage}
}
\subfigure[N=10, $F_0=1000$]{
\begin{minipage}[b]{0.4\linewidth}
\includegraphics[width=1\textwidth]{ZhiyuPictures/N=10_single_particle_energy_F0=1000_s_400_300_Font12.eps} 
\label{fig:thermalization7}
\end{minipage}
}
\caption{Time evolution of the particles' energy $\epsilon$}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

>>>>

With interactions, a term $F_0(1-|r|)\theta(1-|r|)$ is added, where $r$ is the relative
displacement; as shown in Figure \ref{fig:thermalization2}.  Now, the relative motion is no longer
harmonic.  When the internal energy is much larger than $F_0$, one could regard the resulting motion
has having a slightly different frequency, or a collection of frequencies whose center is shifted
slightly from $\omega_0$.  Since the center-of-mass motion is still of frequency $\omega_0$, we have
a superposition of slightly different frequencies, resulting in beating dynamics.  This is clearly
seen in the time evolution of individual energies shown in Figure \ref{fig:thermalization2}.  The
two panels correspond to different internal energy $E_i$ (defined as the total energy minus the
center-of-mass energy).  For larger $E_i$, the distortion of the effective potential (at constant
$F_0$) plays a smaller role in shifting the effective frequency of relative motion; hence the beat
frequency is smaller.

This illustrates a simple mechanism hindering the redistribution of energy between particles, which
is necessary for thermalization or relaxation.  As seen in the example dynamics shown in Figure
\ref{fig:thermalization2}, the difference in energy between the two particles is sustained over
time.  Of course thermalization is not expected anyway in a two-particle system, but we will see
below how this basic effect continues to play a role for larger $N$.

@@

Once we have more than two particles, we expect chaotic or ergodic behavior.  However, it is easy to
imagine mechanisms which drastically slow down redistribution of energy.

Generally, whenever we have some particles with energy very different from others, the dynamics
occurs mostly independently within particle groups of different energies.  For example, considering
a situation with three particles, consider the situation where two of them, say A and B, have small
internal energy, which means their mutual distance and relative velocity are both small, whereas the
third particle (C) has some energy quite different from A and B.
%
Then, A and B will often be interacting, and if their internal energy is smaller than $F_0$,
significant energy exchange can occur between them.  C will generally exchange little energy with
the pair during interactions, in which (due to larger internal energy between C and either A or B),
the interaction is not effective in energy exchange.
%
An example is shown in Figure \ref{fig:thermalization4}, showing the dynamics of $N=5$ particles.
Two pairs of particles persist in performing `internal' dynamics.  The lone particle participates
only in some slow beating motion with the center of mass of one of the pairs.  Clearly, the energy
mismatch acts as a hurdle to the relaxation process.

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{ZhiyuPictures/pair1_pre.eps}
\caption{Time evolution of single-particle energies in a system of five particles, showing a case of
  inefficient energy redistribution.  The energies of the two partiles with highest energy show a beating
  similar to the two-particle case.}
\label{fig:thermalization4}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

%% The argument above still holds in many-particle case. Once we start from some configuration where n
%% particles have a set of close energy levels $\left\lbrace E_n\right\rbrace $ and another m particles
%% have another set of energy levels close to each other $\left\lbrace E_m\right\rbrace $, and
%% $\left\lbrace E_n\right\rbrace $ is quite different from $\left\lbrace E_n\right\rbrace $, we will
%% find these two systems oscillating on the energy level diagram at low frequency.  The discussion
%% above gives us some pictures about the non-ergodic state. For these state, life time is rather long
%% so that once the system reach such configuration (or certain energy distribution), it will take a
%% long time(more than $10^4$ periods for N=5 case) to decay.

\subsection{Condition for relaxation}

The main obstruct to get an thermalized state is the low frequency beat mentioned before. As a
result, to achieve ergodic state, we have to avoid such low frequency oscillation. According to our
discussion before, the solution is to make internal energy of each two-particle pair not ``too
large". Though it is impossible to express the internal energy of every pair in terms of the total
energy $E$, we can estimate it by the average energy, which is $\frac{E}{N}$. At least they are of
the same order. The thermalization condition could be given by:
\begin{equation}
F_0\sigma\sim\frac{E}{N}
\label{eq:thermalizatiton condition}
\end{equation}

%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
\begin{figure}[tbhp]
\centering
%\includegraphics[scale=0.2]{ZhiyuPictures/distribution1_scanF0_E=100_3.eps}
%\caption{Distribution scan $F_0$, $E=100$, $N=5$, $\sigma=1$ {\color{red}{Run again!}} }
% \includegraphics[width=0.32\textwidth]{ZhiyuPictures/N=5_energydistribution_500_400_Font16.eps} 
% \includegraphics[width=0.32\textwidth]{ZhiyuPictures/N=10_energydistribution_500_400_Font16.eps}
% \includegraphics[width=0.32\textwidth]{ZhiyuPictures/N=20_energydistribution_500_400_Font16.eps}
%
\includegraphics[width=0.85\textwidth]{ZhiyuPictures/Boltzmann_2panels_01-crop.pdf}
%
\caption{the energy distribution measured in $\sim 10000$ time units, as will be verified later, Green curves corresponds to Boltzmann distribution; Red curves are obviously non-Boltzmann, while the yellow ones are intermediate case, represent ``thermalization threshold". In three figures for different N, we choose E/N identical (=1000), and thermalization is reached at $F_0=500\sim1000$, which is consistent with our prediction in Eq.\ref{eq:thermalizatiton condition}}
\label{fig:thermalization5}
\end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%

@@

%% %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% 
%% \begin{figure}[h]
%% \subfigure[N=10, $F_0=200$]{
%% \begin{minipage}[b]{0.4\linewidth}
%% \includegraphics[width=1\textwidth]{ZhiyuPictures/N=10_single_particle_energy_F0=200_s_400_300_Font12.eps}
%% \label{fig:thermalization6}
%% \end{minipage}
%% }
%% \subfigure[N=10, $F_0=1000$]{
%% \begin{minipage}[b]{0.4\linewidth}
%% \includegraphics[width=1\textwidth]{ZhiyuPictures/N=10_single_particle_energy_F0=1000_s_400_300_Font12.eps} 
%% \label{fig:thermalization7}
%% \end{minipage}
%% }
%% \caption{Time evolution of the particles' energy $\epsilon$}
%% \end{figure}
%%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%% FIGURE %%%%%
